---
title: "Data Exploration"
author: "Aubrey Ahern"
date: "11/8/2022"
output: html_document
---

set things up with the typical packages
```{r}
rm(list = ls())
library(here)
library(ggplot2)
library(dplyr)
```

suck in data
```{r}
DF <- read.csv(here("Data", "fixed-deaths.csv"), stringsAsFactors = TRUE)
DF$count <- as.numeric(DF$count)
#changed count data to numeric now, so I don't have to later
#all other variables should be categorical so stringsAsFactors should work fine for those
```

check for mistakes in my data

age ranges
```{r}
levels(DF$age_at_death)
```
genders
```{r}
levels(DF$current_gender)
```
races
```{r}
levels(DF$race_ethnicity)
```
transmission categories
```{r}
levels(DF$transmission_category)
```

no errors that i can see... let's explore



## Relationship #1
let's first see if there is any notable relationship between gender versus cause deaths by HIV/AIDS
- create a data frame containing only the data i want 
```{r}
DF_gender <- DF %>%
  group_by(current_gender) %>%
  summarise(count = sum(count)) %>%
  arrange(desc(count)) %>%
  na.omit()

DF_gender
```

- plot it
```{r}
ggplot(DF_gender, aes(current_gender, count))+
     geom_col()
```

- gender is a categorical variable and cause deaths by HIV/AIDS is a numeric variable
- from this plot we can see that primarily trans and cis women suffer from cause deaths by HIV/AIDS
- the data ranges from genders with cause deaths less than 25 to almost 400



## Relationship #2
now lets see if there is any relationship between HIV cause deaths for trans and cis women over time

- create a data frame for trans women
```{r, results = 'hide'}
DF_transwomen <- DF %>%
  filter(current_gender == "Transgender women")%>%
  mutate(year = as.numeric(year),
         count = as.numeric(count))
DF_transwomen
```

- create a data frame for cis women
```{r, results = 'hide'}
DF_ciswomen <- DF %>%
  filter(current_gender == "Cisgender women")%>%
  mutate(year = as.numeric(year),
         count = as.numeric(count))
DF_ciswomen
```

- put them both into a graph to visualize them
```{r}
ggplot(data = DF_ciswomen, aes(x = year, y = count, colour = current_gender)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(data = DF_transwomen, aes(x = year, y = count, color = current_gender)) +
  geom_smooth(data = DF_transwomen, aes(x = year, y = count, color = current_gender), method = "lm", se = FALSE)
```

- year, in this case is considered to be a continuous variable and cause deaths by HIV/AIDS is considered to be a numeric variable
- from this plot we can see that generally, cause deaths for ciswomen decrease over time while cause deaths for transwomen increase over time
- the cause deaths by HIV/AIDS for cis and trans women ranges from 40 to over 60



## Relationship #3
lastly, let's see if there's any relationship between race and cause deaths by HIV/AIDS

- create a data frame containing only the data I want
```{r}
DF_race_ethnicity <- DF %>%
  group_by(race_ethnicity) %>%
  summarise(count = sum(count)) %>%
  arrange(desc(count)) %>%
  na.omit()

DF_race_ethnicity
```

- plot it
```{r}
ggplot(DF_race_ethnicity, aes(race_ethnicity, count))+
     geom_col()+
  labs(x = "Race / Ethnicity",
       y = "Cause Deaths by HIV/AIDS")
```

- race/ethnicity is a categorical variable while cause deaths by HIV/AIDS is considered to be a numeric variable
- from this plot, we can see that generally, white people experienced more cause deaths by HIV/AIDS
- cause deaths by HIV/AIDS range from a little less than 250 for Hawaiian/Pacific Islander people to nearly 1000 for White people

- These are the three relationships that I will explore further in my data analysis draft. 
