---
title: "Cap DAP Final Draft"
author: "Aubrey Ahern"
date: "12/14/2022"
---



## Introduction

- I took my data from https://data.chhs.ca.gov/dataset/hiv-aids-cases. 
- The data was collected from 2011-2017 within the state of California in order to collect public health information. 
- The data outlines all cause deaths in HIV/AIDS by gender, age, race/ethnicity, and transmission category. 
- I used this data to take a look at the distribution of HIV/AIDS cause deaths by different demographic variables, primarily looking at how good of a predictor gender is in determining health outcomes. 
- While there is no biologic reasoning for why certain groups of people experience HIV/AIDS cause deaths at a higher rate, we can call upon public health knowledge to answer these questions.
  - When taking social determinants of health into consideration, we may see that women and LGBTQ+ communities experience lower accessibility to health care and are often subjected to discrimination within the health care field. This may provide an explanation to the data that will be presented throughout this project (Sharman). 

 
 
#### Getting Started

- I began by setting things up as usual and importing the typical packages, *here*, *ggfortify*, and *tidyverse* . I also had to import the *AICmodavg* package in order to one run of my statistical tests later on. 

```{r, include = FALSE} 
rm(list = ls())
library(here)
library(ggfortify)
library(tidyverse)
library(AICcmodavg)
```
 
 
- I made sure my data was compatible to work with R by...
    - Excluding spaces and special characters from column names 
    - Organizing categories into columns instead of rows
    - Making sure that each cell contained only one piece of information. 
- After importing my data, I made all variables factors, except for count which I made numeric.

#### Limitations

- While this data set contains a lot of information on demographic information on those who have died from HIV/AIDS, it only includes **one** piece of information per row, instead of multiple per datum. I have worked with the data as best I could but would potentially use a more extensive data set in the future. 
 
 
```{r, include = FALSE}
DF <- read.csv(here("Data", "fixed-deaths.csv"), stringsAsFactors = TRUE)
DF$count <- as.numeric(DF$count)
```



## Analysis
### Test #1

```{r, results = 'hide'}
DF_gender <- DF %>%
  group_by(current_gender) %>%
  summarise(count = sum(count)) %>%
  arrange(desc(count)) %>%
  na.omit()

DF_gender
```


- First, let's visualize this relationship.
```{r}
ggplot(DF_gender, aes(current_gender, count))+
     geom_col()
```

- Now I can use my visualization to formulate my hypotheses. 

#### Hypothesis: *Trans women and cis women experience HIV/AIDS cause deaths at a higher rate than those of other genders. *
#### Null Hypothesis: *There is no correlation between gender and HIV/AIDS cause deaths.*

- I then ran a statistical test to see if I can accept or reject it. This relationship includes one piece of categorical data (current gender) and one piece of count data (count). Thus, I will conduct an ANOVA test.

```{r}
DF_model <- DF %>%
  slice(106:140)
myMod1 <- lm(count ~ current_gender, data = DF_model)
anova(myMod1)
```
- Trans women and cis women experienced significantly more deaths by HIV-AIDS than those of other genders as the p value was 5.778x10^-16^ which is less than 0.05. 
   
   
### Test #2  

- Since we now know that trans and cis women experience more deaths by HIV-AIDS than those of other genders, let's analyze the relationship between year and deaths for trans and cis women. 

- I created a new data frames that only contains data for trans women.
```{r, results = 'hide'}
DF_transwomen <- DF %>%
  filter(current_gender == "Transgender women")%>%
  mutate(year = as.numeric(year),
         count = as.numeric(count))
DF_transwomen
```

- Then I visualized this new dataframe.
```{r}
ggplot(DF_transwomen, aes(year, count))+
     geom_col()
```

- This allowed me to form my hypotheses:
##### Hypothesis: *Cause deaths by HIV/AIDS increase over time for transwomen*
##### Null Hypothesis: *There is no correlation between HIV/AIDS cause deaths for trans women by year.*

- Then, I ran a t-test to see if there is a relationship between time (in years) and death count for trans women.
    - I chose to conduct a t-test rather than another ANOVA test as I considered year to be numeric data rather than categorical. 
```{r}
myMod2 <- lm(count ~ year, data = DF_transwomen)
summary(myMod2)
```

- This allowed me to accept the null hypothesis that there was no observable relationship between year and death count for trans women as the p value, 0.2964 > 0.05. Below is a figure which represents the (lack of) correlation within this data.

```{r, results = 'hide'}
ggplot(DF_transwomen, aes(x = year, y = count)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)+
  labs(x = "Year",
       y= "HIV/AIDS Cause Deaths",
       title = "Trans Women Deaths by Year") +
  theme_minimal()
```

- I then did the same for cis women by first creating a new data frame that only contains the data I wish to analyze.
```{r, results = 'hide'}
DF_ciswomen <- DF %>%
  filter(current_gender == "Cisgender women")%>%
  mutate(year = as.numeric(year),
         count = as.numeric(count))
DF_ciswomen
```

- Again, I visualized this new data frame...
```{r}
ggplot(DF_ciswomen, aes(year, count))+
     geom_col()
```

- This allowed me to form my hypotheses:
##### Hypothesis: *Cause deaths by HIV/AIDS decrease over time for cis women.*
##### Null Hypothesis: *There is no correlation between HIV/AIDS cause deaths for cis women by year.*

```{r}
myMod3 <- lm(count ~ year, data = DF_ciswomen)
summary(myMod3)
```

- This allowed me to accept the null hypothesis that there was no observable relationship between year and death count for cis women as the p value, 0.1858 > 0.05. Below is a figure which represents the (lack of) correlation within this data.

```{r}
ggplot(DF_ciswomen, aes(x = year, y = count)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)+
  labs(x = "Year",
       y= "HIV/AIDS Cause Deaths",
       title = "Cis Women Deaths by Year") +
  theme_minimal()
```

### Test #3
##### Hypothesis: *Gender is the best predictor for HIV/AIDS cause deaths.*
##### Null Hypothesis: *Gender is not the best predictor for HIV/AIDS cause deaths.*

- Lastly, I considered which category, race, gender, age, or transmission method, is the best predictor of deth count by HIV/AIDS. 
- This calls for a new test called an AIC test. An Akaike information criterion, or an AIC test, allows one to assess how well different models fit a set of data and determine which factor is the best predictor of the data.

For this test, I first had to make linear models for each category...
```{r}
mod_race <- lm(count ~ race_ethnicity, data = DF)
mod_gender <- lm(count ~ current_gender, data = DF)
mod_age <- lm(count ~ age_at_death, data = DF)
mod_transmission <- lm(count ~ transmission_category, data = DF)

death_models <- list(mod_race, mod_gender, mod_age, mod_transmission)
names(death_models) <- c("Race", "Gender", "Age", "Transmission")
```

```{r}
summary(death_models)
```


Then I was able to run the test...
```{r}
AICdeaths <- aictab(cand.set = death_models, second.ord = TRUE, sort = TRUE)
AICdeaths
```
I had to refresh myself on how to read an AIC table and outlined what key pieces of information within the table mean below:
- k = the number of parameters in the model
- AICc = the information score of the model (the smaller the AIC value the better the model fit)
- Delta_AICc = tells you how much worse a variable is than the last predictor listed

Based on this table, we can see that gender is in fact the best predictor for cause deaths by HIV/AIDS. 

## Biological Summary
- There is no biological reasoning for why certain genders experience HIV/AIDS at higher rates than others aside from the fact that the virus is present in their communities. 
- As seen throughout the analysis portion of this project, gender is the best categorical predictor of HIV/AIDS cause deaths but this does not mean in any way that gender causes HIV/AIDS. Instead, the current social order discriminates against certain groups such as trans folks and women based on their gender. The medical field is no exception from this social order and thus, these groups struggle to access equitable health care and therefore experience worse health outcomes (Sharman). 
- People who are within marginalized communities often experience instances where physicians minimize their pain and ignore important health indicators such as the ones observed in HIV/AIDS. This can lead to misdiagnoses as well as late diagnoses (Kronenfeld). Catching HIV/AIDS within its early stages is crucial in dictating how the disease progresses. Thus, this may be a reason why trans women and cis women are more likely to experience HIV/AIDS cause deaths. 

## Challenges
- The main take away from this analysis is that gender is a key predictor for HIV/AIDS cause deaths and that trans and cis women are more likely to experience death due to HIV/AIDS.
- After completing this project, there is a few things I would do differently...
    - I would spend more time considering which data I would use. The data I found was pretty difficult to work with and didn't contain as much information as I had hoped. 
    - I would find data with a larger sample size over a greater amount of time. 
        - Certain analyses could have been improved if my data spanned over more years.
    - I would conduct more tests if possible.
    
## References
Zena Sharman. (2016). The Remedy : Queer and Trans Voices on Health and Health Care. Arsenal Pulp Press.

Jennie Jacobs Kronenfeld. (2018). Gender, Women’s Health Care Concerns and Other Social Factors in Health and Health Care: Vol. First edition. Emerald Publishing Limited

