---
title: "CAP DAP Presentation"
author: "Aubrey Ahern"
date: "11/30/2022"
---



## Introduction

- I took my data from https://data.chhs.ca.gov/dataset/hiv-aids-cases. 
- The data outlines all cause deaths in HIV/AIDS by gender, age, race/ethnicity, and transmission category. 
- The data was collected from 2011-2017 within the state of California in order to collect public health information. 
 
 
 
## Getting Started

- I began by setting things up as usual and importing the typical packages, *here*, *ggfortify*, and *tidyverse* . I also had to import the *AICmodavg* package in order to one run of my statistical tests later on. 

```{r, include = FALSE} 
rm(list = ls())
library(here)
library(ggfortify)
library(tidyverse)
library(AICcmodavg)
```
 
 
- I made sure my data was compatible to work with R by...
    - Excluding spaces and special characters from column names 
    - Organizing categories into columns instead of rows
    - Making sure that each cell contained only one piece of information. 
- After importing my data, I made all variables factors, except for count which I made numeric.

## Limitations

- While this data set contains a lot of information on demographic information on those who have died from HIV/AIDS, it only includes **one** piece of information per row, instead of multiple per datum. I have worked with the data as best I could but would potentially use a more extensive data set in the future. 
 
 
```{r, include = FALSE}
DF <- read.csv(here("Data", "fixed-deaths.csv"), stringsAsFactors = TRUE)
DF$count <- as.numeric(DF$count)
```



## Analysis
### Test #1
##### Hypothesis: *There is a correlation between gender and HIV/AIDS cause deaths.*
##### Null Hypothesis: *There is no correlation between gender and HIV/AIDS cause deaths.*

```{r, results = 'hide'}
DF_gender <- DF %>%
  group_by(current_gender) %>%
  summarise(count = sum(count)) %>%
  arrange(desc(count)) %>%
  na.omit()

DF_gender
```


- First, let's visualize this relationship.
```{r}
ggplot(DF_gender, aes(current_gender, count))+
     geom_col()
```

- This initial visualization supported my hypothesis. I then ran a statistical test to see if I can accept or reject it. This relationship includes one piece of categorical data (current gender) and one piece of count data (count). Thus, I will conduct an ANOVA test.

```{r}
DF_model <- DF %>%
  slice(106:140)
myMod1 <- lm(count ~ current_gender, data = DF_model)
anova(myMod1)
```
- Trans women and cis women experienced significantly more deaths by HIV-AIDS than those of other genders as the p value was 5.778x10^-16^ which is less than 0.05. 
   
### Test #2  
##### Hypothesis: *There is a correlation between HIV/AIDS cause deaths for trans and cis women by year.*
##### Null Hypothesis: *There is no correlation between HIV/AIDS cause deaths for trans and cis women by year.*

- Since we now know that trans and cis women experience more deaths by HIV-AIDS than those of other genders, let's analyze the relationship between year and deaths for trans and cis women. 

- I created a new data frames that only contains data for trans women.
```{r, results = 'hide'}
DF_transwomen <- DF %>%
  filter(current_gender == "Transgender women")%>%
  mutate(year = as.numeric(year),
         count = as.numeric(count))
DF_transwomen
```

- Then, I ran a t-test to see if there is a relationship between time (in years) and death count for trans women.
    - I chose to conduct a t-test rather than another ANOVA test as I considered year to be numeric data rather than categorical. 
```{r}
myMod2 <- lm(count ~ year, data = DF_transwomen)
summary(myMod2)
```

- This allowed me to accept the null hypothesis that there was no observable relationship between year and death count for trans women as the p value, 0.2964 > 0.05. Below is a figure which represents the (lack of) correlation within this data.

```{r, results = 'hide'}
ggplot(DF_transwomen, aes(x = year, y = count)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)+
  labs(x = "Year",
       y= "HIV/AIDS Cause Deaths",
       title = "Trans Women Deaths by Year") +
  theme_minimal()
```

- I then did the same for cis women by first creating a new data frame that only contains the data I wish to analyze.
```{r, results = 'hide'}
DF_ciswomen <- DF %>%
  filter(current_gender == "Cisgender women")%>%
  mutate(year = as.numeric(year),
         count = as.numeric(count))
DF_ciswomen
```


```{r}
myMod3 <- lm(count ~ year, data = DF_ciswomen)
summary(myMod3)
```

- This allowed me to accept the null hypothesis that there was no observable relationship between year and death count for cis women as the p value, 0.1858 > 0.05. Below is a figure which represents the (lack of) correlation within this data.

```{r}
ggplot(DF_ciswomen, aes(x = year, y = count)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)+
  labs(x = "Year",
       y= "HIV/AIDS Cause Deaths",
       title = "Cis Women Deaths by Year") +
  theme_minimal()
```

### Test #3
##### Hypothesis: *Gender is the best predictor for HIV/AIDS cause deaths.*
##### Null Hypothesis: *Gender is not the best predictor for HIV/AIDS cause deaths.*

- Lastly, I considered which category, race, gender, age, or transmission method, is the best predictor of deth count by HIV/AIDS. 
- This calls for a new test called an AIC test. An Akaike information criterion, or an AIC test, allows one to assess how well different models fit a set of data and determine which factor is the best predictor of the data.

For this test, I first had to make linear models for each category...
```{r}
mod_race <- lm(count ~ race_ethnicity, data = DF)
mod_gender <- lm(count ~ current_gender, data = DF)
mod_age <- lm(count ~ age_at_death, data = DF)
mod_transmission <- lm(count ~ transmission_category, data = DF)

death_models <- list(mod_race, mod_gender, mod_age, mod_transmission)
names(death_models) <- c("Race", "Gender", "Age", "Transmission")
```

Then I was able to run the test...
```{r}
AICdeaths <- aictab(cand.set = death_models, second.ord = TRUE, sort = TRUE)
AICdeaths
```
I had to refresh myself on how to read an AIC table and outlined what key pieces of information within the table mean below:
- k = the number of parameters in the model
- AICc = the information score of the model (the smaller the AIC value the better the model fit)
- Delta_AICc = tells you how much worse a variable is than the last predictor listed

Based on this table, we can see that gender is in fact the best predictor for cause deaths by HIV/AIDS. 


## Conclusion
- The main take away from this analysis is that gender is a key predictor for HIV/AIDS cause deaths and that trans and cis women are more likely to experience death due to HIV/AIDS.

## Reflection
- After completing this project, there is a few things I would do differently...
    - I would spend more time considering which data I would use. The data I found was pretty difficult to work with and didn't contain as much information as I had hoped. 
    - I would find data with a larger sample size over a greater amount of time. 
        - Certain analyses could have been improved if my data spanned over more years.
    - I would conduct more tests if possible.
    


